---
title: "Analysis of Variance"
author: "Mwangi George"
date: "2022-10-27"
output: 
  github_document:
    toc: yes
---

# Analysis of variance with the `gapminder` dataset
```{r include=FALSE}
pacman::p_load(tidyverse, ggthemes, gapminder)
```
If we wanted to test if there exist a significant difference in the GDP per Capita between two continents in the gapminder dataset, we would utilize a t-test. However, if we wanted to test if there exist a significant difference in the GDP per Capita in more than two continent, Anova is our go-to test. In a nutshell, Anova is used when comparing group means between more than two groups.

## Example Question

Is the difference between the GDP per Capita of all continents in the year 2007 significant at 5% level of significance?

*Before answering this question, we have to wrangle the data into a usable format*
```{r}
# data Wrangling
gap_2007 <- gapminder %>% 
  # filter for rows that fit the question
  filter(year == 2007) %>% 
  # select variables of interest
  select(continent, gdpPercap)

# print the first 6 rows
head(gap_2007)
```

# Exploratory Analysis 

Before conducting the test it is important to do some exploratory analysis, that is, calculate the summary statistic for each group and visualize the distribution of the numeric variable for each group using boxplots.
```{r}
gap_2007 %>% 
  # groups the results by continent
  group_by(continent) %>% 
  # calculate summaries for the above groups
  summarise(mean_gdpPercap = mean(gdpPercap))
```

By just looking at the above figures, we can see a big difference in the mean GDP per Capita of the three continents in the year 2007. However, we can't stop at that point; we have to test if the difference is statistically significant since it could be caused by chance or sampling variability. Lets visualize the distributions
```{r fig.width=10}
gap_2007 %>% 
  # base plot layer
  ggplot(aes(continent, gdpPercap, color = continent))+
  # geometric representation layer
  geom_boxplot()+
  # theme layer
  theme_economist()+
  theme(legend.position = "none")+
  # labeling
  labs(title = "Distributions of GDP Per Captita for all continents",
       subtitle = "(Source::gapminder library)")
```

The box plots made it look like the distribution of gdpPercap was different for each of the 5 continents. However, it didn't tell us whether the mean gdpPercap was different in each continent group. To determine that, we can use an ANOVA test. 

The null and alternative hypotheses can be written as follows:

* *H1: U1 = U2 = U3 = U4 = U5*
* *H2: U1 != U2 != U3 != U4 != U5*

We call the function `lm()` to fit a linear regression using gdpPercap as Y variable and `continent` as predictor variable
```{r}
# fit linear regression
gap_2007_model <- lm(gdpPercap ~ continent, data = gap_2007)
```

Next, we call the `anova()` on the model to perform an analysis of variance
```{r}
# remove scientific notation
options(scipen = 999)

# perform Anova
anova(gap_2007_model)
```

The p_value (0.000000000000001127) is obviously less than our alpha value (0.05).
We reject the null hypothesis (H1) in favor of the alternative hypothesis that there exist a significance difference in the GDP per Capita of the 5 continents in 2007. 

# Limitation of `anova()`

The problem with the above method is that it does not show us the p_values within pairs. We should be able to know if there exists a statistically significance difference in any of the continent pairs. 

Therefore, we use the pairwise t.test to test each pair at 5 percent level of significance. The results show a matrix of 10 p_values from each possible pair. 
```{r}
# remove scientific notation
options(scipen = 999)

# pairwise t test
# the 1st argument is the numeric variable whose mean we are interested in
# the 2nd argument is the categorical variable defining the groups
pairwise.t.test(gap_2007$gdpPercap, 
                gap_2007$continent,
                p.adjust.method = "none")
```

From the output above, we get two P_values that are greater than our alpha value(0.05). We are are confident that, at 5% level of significance, the differences in the mean gdpPercap between Americas and Asia, as well as that of Europe and Oceania are not statistically significant. 

# Adjustment

In the above test, we set `p.adjust.method` to `none`. Generally, as the number of groups/categories increases, the number of pairs and hence the number of hypothesis tests increase quadratically. The more tests we run, the higher the probability that at least one of them will give a false positive result (Type I error). This means that we are more likely to reject the null hypothesis while it is actually true in one of the tests. 

The solution to these problem is to apply an adjustment to increase the P_values, thereby reducing the probability of getting a false positive. R provides several adjustment methods as shown by the code below.
```{r}
p.adjust.methods
```

Let's see the output of our pairwise t test when we set `p.adjust.method` to `holm`.
```{r}
# remove scientific notation
options(scipen = 999)

# pairwise t test setting adjustment to holm
pairwise.t.test(gap_2007$gdpPercap, 
                gap_2007$continent,
                p.adjust.method = "holm")
```

The output reveals an increase in the p_values that are greater than 0.05 from 2 to 3. Between pairs Americas and Asia, Asia and Oceania, and Europe and Oceania, we fail to reject the null hypothesis and conclude that the difference in the mean GDP per Capita is not statistically significant between these pairs. It could be caused by chance. 


### Contact Me

[LinkedIn](https://www.linkedin.com/in/georgemwangikenya)
[Twitter](https://twitter.com/mwangi__george)
