---
title: "Shiny Fundamentals"
author: "Mwangi George"
date: "2022-12-27"
output: html_document
---

```{r}
pacman::p_load(shiny, tidyverse, babynames)

# create a user interface
ui <- fluidPage(
  "Hello, World!!"
)

# define a custom function to create the server
server <- function(input, output, session) {

}

# Run the app
shinyApp(ui = ui, server = server)


# Ask a question with an input
ui <- fluidPage(
  textInput(
    "name",
    "Enter a name:"
  ),
  selectInput(
    inputId = "Drink_type",
    label = "Tea or Coffee",
    choices = c("Tea", "Coffee")
  ),
  textOutput("q"),
  textOutput("drink")
)

server <- function(input, output) {
  output$q <- renderText({
    paste(
      "Do you prefer tea or coffee,", input$name, "?"
    )
  })
  output$drink <- renderText({
    paste(
      "I prefer", input$Drink_type, "!"
    )
  })
}

shinyApp(ui, server)

# Another one
ui <- fluidPage(
  textInput("name", "What is your name?"),
  textOutput("greeting")
)

server <- function(input, output) {
  output$greeting <- renderText({
    paste("Hi,", input$name)
  })
}

shinyApp(ui, server)


# Building a babynames explorer
library(babynames)
library(tidyverse)
head(babynames)
# create the ui
ui <- fluidPage(
  titlePanel("Baby Name Explorer"),
  sidebarLayout(
    sidebarPanel(
      textInput("name", "Enter Name"),
      sliderInput("year_of_birth", "Select Year", 1880, 2017, 2010)
    ), 
    mainPanel(
      plotOutput("trend")
    )
  )
)

# Server
server <- function(input, output, session) {
  
  output$trend <- renderPlot({
    ggplot(
     data = subset(babynames, name == input$name, year = input$year_of_birth)
      ) +
      geom_line(
        aes(
          x = year,
          y = prop,
          color = sex
        )
      ) +
      labs(
        title = "Popularity of a name over time",
        y = "proportion",
        caption = "Made by George Ngugi",
        subtitle = "Analysis of full baby name data provided by the SSA"
      ) +
      ggthemes::theme_economist()
  })
}

# run the app
shinyApp(ui, server)


# create the user interface
ui <- fluidPage(
  titlePanel("Baby Names Explorer"),
  # CODE BELOW: Add select input named "sex" to choose between "M" and "F"
  sidebarLayout(
  sidebarPanel(
    selectInput(
      "sex", "Choose Sex", choices = c("M", "F"), selected = "F"
    ),
    sliderInput(
      "year",
      "select year",
      min = 1900,
      max = 2017,
      value = 2017
      )
  ),
  # Add plot output to display top 10 most popular names
  mainPanel(
    plotOutput('plot_top_10_names')
    )
  )
)

server <- function(input, output, session) {
  # Render plot of top 10 most popular names
  output$plot_top_10_names <- renderPlot({
    # Get top 10 names by sex and year
    top_10_names <- babynames %>%
      # MODIFY CODE BELOW: Filter for the selected sex
      filter(sex == input$sex) %>%
      filter(year == input$year) %>%
      slice_max(prop, n = 10)
    # Plot top 10 names by sex and year
    ggplot(
      data = top_10_names, 
      aes(
        x = fct_reorder(name, prop), 
        y = prop
        )
      ) +
      geom_col(fill = "#263e63") +
      coord_flip() +
      labs(
        title = "Top 10 Names in Selected Year",
        caption = "Data Source::Babynames Package",
        x = "",
        y = "Popularity"
      ) +
      ggthemes::theme_few() +
      theme(
        axis.text = element_text(
          face = "bold",
          size = 16
        ),
        plot.title = element_text(
          face = "bold", 
          size = 17
          ),
        axis.title.y = element_text(
          face = "bold",
          size = 16
        ),
        plot.caption = element_text(
          face = "bold",
          size = 16
        )
      )
  })
}

shinyApp(ui, server)

```


# Other Input Functions 

You can use a sliderInput to allow users to select a year. A selectInput is a great way to allow for a selection from a list of fixed options, such as a preference for dogs or cats. The numericalInput allows you to provide a range of numbers users can choose from, which they can increase or decrease using the little arrows. A dateRangeInput allows you to provide users with a set of dates, and a calendar drop down appears when they click so they can select a specific one. These are just four examples, and Shiny definitely provides more.
All input functions have their first argument, an inputId, in common. The inputId needs to be a character string, and each input should have a unique id so you can refer to it in the server to make updates to the app. Many inputs have a label as their next argument, which is a character string that is often shown to the user to let them know what they should do with the input. From there, each input function has unique arguments that help you successfully build the app. A selectInput requires a list of choices. The user will automatically see the first choice in the list. A sliderInput requires value that the slider will be set at by default, then a min and max of the other values users can choose from. You'll always use input functions in the UI of your app. 


# Render functions
Render functions are used to build outputs in the server based on inputs and possibly other things, like other parts of a character string, as in this example. Recall our question and answer app, which you saw the code for in the last lesson. This app outputs two different character strings, one with a question and one with the answer. Given they're both character strings, it's appropriate to use the renderText function to create both outputs.

# Other render functions
Like inputs, Shiny provides a number of other render functions you can use to create a number of different kinds of outputs beyond just text, including rendertable, renderImage, renderPlot, and more. The Shiny documentation is a great place to go to find out about more render functions available to you.

# Output functions
Output functions are used back in the UI to display the outputs built in the server with render functions. For our question and answer app, we used two textOutput functions to display both the question and the answer outputs.

# Other output functions
Like inputs and render functions, Shiny provides a wide variety of outputs depending on what kind of output you've built, whether that be a table, an image, or a plot.

# Non-Shiny output and render functions
There are packages outside of Shiny that provide ways to build outputs with render and output functions. Thanks to htmlwidgets, packages such as DT, leaflet, and plotly allow you to build interactive data tables, maps, and plots as Shiny outputs. For example, the DT package allows you to build interactive data tables versus static ones. This code, for example, will build an app creating an interactive data table version of a random 10% of the babynames dataset we've been using throughout the course. Notice that we use the renderDT function to build the output in the server, then the DTOutput function in the ui to display the table, which we called babynames_table. The interactive table is sortable, searchable, plus users can choose to show 10, 25, or more entries at a time.
```{r}
# install.packages("DT")
ui <- fluidPage(
  titlePanel("Baby Names Explorer"),
  DT::DTOutput("babynames_table")
)

server <- function(input, output){
  output$babynames_table <- DT::renderDT({
    babynames %>% 
      slice_sample(prop = .1) %>% 
      DT::datatable()
  })
}

shinyApp(ui, server)
```


In order to add any output to a Shiny app, you need to:

* Create the output (plot, table, text, etc.).
    
* Render the output object using the appropriate render___ function.
    
* Assign the rendered object to output$x.
    
* Add the output to the UI using the appropriate ___Output function.
```{r}
ui <- fluidPage(
  titlePanel("What's in a Name?"),
  # Add select input named "sex" to choose between "M" and "F"
  sidebarLayout(
    sidebarPanel(
  selectInput('sex', 'Select Sex', choices = c("F", "M")),
  # Add slider input named "year" to select year between 1900 and 2010
  sliderInput('year', 'Select Year', min = 1900, max = 2010, value = 1900)),
  # CODE BELOW: Add table output named "table_top_10_names"
  mainPanel(
  tableOutput("table_top_10_names")
  )
  )
)

server <- function(input, output, session){
  # Function to create a data frame of top 10 names by sex and year 
  top_10_names <- function(){
    babynames %>% 
      filter(sex == input$sex) %>% 
      filter(year == input$year) %>% 
      slice_max(prop, n = 10)
  }
  # CODE BELOW: Render a table output named "table_top_10_names"
  output$table_top_10_names <- renderTable({
    top_10_names()
  })
  
}

shinyApp(ui = ui, server = server)
```


There are multiple htmlwidgets packages like DT, leaflet, plotly, etc. that provide highly interactive outputs and can be easily integrated into Shiny apps using almost the same pattern. For example, you can turn a static table in a Shiny app into an interactive table using the DT package:

Create an interactive table using DT::datatable().

Render it using DT::renderDT().

Display it using DT::DTOutput().

```{r}
# Create the user interface 
ui <- fluidPage(
  titlePanel("Baby Names Explorer"),
  selectInput("sex", "Select Sex", 
              choices = c("M", "F"), selected = "M"),
  sliderInput("year", "Select Year", min = 1880, max = 2017, value = 2017),
  DT::DTOutput("table_top_10_names")
)

# Create the Server 
server <- function(input, output, session){
  top_10_names <- function(){
    babynames %>% 
      filter(sex == input$sex) %>% 
      filter(year == input$year) %>% 
      slice_max(prop, n = 10)
  }
  output$table_top_10_names <- DT::renderDT({
    top_10_names() %>% 
       DT::datatable()
  })
}

shinyApp(ui = ui, server = server)
```



Similar to creating interactive tables, you can easily turn a static plot created using ggplot2 into an interactive plot using the plotly package. To render an interactive plot, use plotly::renderPlotly(), and display it using plotly::plotlyOutput().


```{r}
# group data by year and select top 10 names in each year
top_trendy_names <- babynames %>% 
  group_by(year) %>% 
  slice_max(prop, n = 10)

# build the user interface
ui <- fluidPage(
  # App title
  titlePanel("Baby Names Explorer"),
  # use the superhero theme
  theme = shinytheme("superhero"),
  # build app layout
  sidebarLayout(
    sidebarPanel(
      selectInput(
        'name', 
        'Select Name', 
        choices = top_trendy_names$name, 
        selected = "Mary"
        ),
      # Add a plotly output named 'plot_trendy_names'
      selectInput(
        "sex", 
        "Select Sex", 
        choices = c("M", "F"), 
        selected = "M"
        ),
      sliderInput(
        "year", 
        "Select Year", 
        min = 1880, 
        max = 2017, 
        value = 2017
        )
      ),
    mainPanel(
      tabsetPanel(
        tabPanel(
          "Chart", 
          plotly::plotlyOutput("plot_trendy_names")
          ),
        tabPanel(
          "table",
          DT::DTOutput("table_top_10_names")
          )
        )
      )
  )
)

# build the server
server <- function(input, output, session){
  # Function to plot trends in a name
  plot_trends <- function(){
     babynames %>% 
      filter(name == input$name) %>% 
      ggplot(aes(x = year, y = n)) +
      geom_col()
  }
  # Render a plotly output named 'plot_trendy_names'
  output$plot_trendy_names<- plotly::renderPlotly({
    plot_trends()
  })
   top_10_names <- function(){
    babynames %>% 
      filter(sex == input$sex) %>% 
      filter(name == input$name) %>% 
      filter(year == input$year) %>% 
      slice_max(prop, n = 10)
  }
  output$table_top_10_names <- DT::renderDT({
    top_10_names() %>% 
       DT::datatable()
  })
}

# Run the App
shinyApp(ui = ui, server = server)
```



2. Default Shiny app layout
The default Shiny app layout, shown here with the default Shiny app example, a histogram of the wait times between eruptions of the geyser Old Faithful at Yellowstone National Park, simply stacks the elements of your app on top of one another, one by one. You can see here that the title is on top of the selector, which is all on top of the actual histogram output. While the default layout might be fine for simple apps, as they start to get more complicated, it would be nice to be able to control where elements of the app end up.

3. Sidebar layout
One of the simplest layouts you can choose is a sidebar layout. You can see that the title is still at the top, justified left, but the slider is now in the sidebar panel at the left, while the histogram is in the main panel to the right. You might recall that you already made use of this layout in Chapter 1. To adjust the layouts, adjust the code in the UI. The code that creates the sliderInput and outputs the plot is overall wrapped in a sidebarLayout function. The slider itself should be put into sidebarPanel, while the plotOutput function should be put into mainPanel. These small adjustments to your code go a long way towards making your app more aesthetically appealing.

4. Tab layout
If you want to display multiple things in your app, such as two plots or a plot and a table, it can be visually noisy to stack them on top of one another in the main panel. Instead, you can consider using a tab layout, like shown here. One tab has the histogram of Old Faithful wait times, while the other shows a histogram of eruptions.

5. Tab layout
Technically, a tab layout is just an extension of a sidebar layout, so that function is still present in the UI, wrapping all of the code, and the sidebar panel also remains unchanged. To add the tab layout, you have to add a tabsetPanel inside of the main panel. Each individual tab has to then be created with tabPanel, and you should give each a label.

6. Theme selector
In addition to layouts, the shinythemes package allows you to make use of prebuilt themes that allow you to change the color scheme of your app. If you're not sure what theme you want to use, you don't have to cycle through them by changing the code. Instead, you can add the themeSelector function to your UI, which will add a dropdown selector in the upper right corner of your app and allow you to see what all of the pre-built themes look like when applied to your app. In this example, you can see the superhero theme applied to the Old Faithful histogram app.

7. Adding a theme
Once you've selected a theme you're interested in using the themeSelector, you can add it to your app using the theme argument in the UI. If you're interested in using the same superhero theme, just add this line, here inserted between the title and the sidebar layout code, which is good Shiny app coding practice. The themeSelector dropdown will now disappear, and the app is hard-coded with the chosen theme.



 Explore Life Expectation vs. GDP per Capita
In this app, users will explore the relationship between life expectancy and GDP per capita. Users will select a continent and a year. Based on their choice, the app will display a scatterplot of life expectation versus GDP per capita.

3. Explore Life Expectation vs. GDP per Capita
A tab layout will allow switching between the plot and an interactive datatable. The table will give users the opportunity to explore the raw data used to build the plot.

4. Building Shiny apps: 4 steps
There are many different ways to build a Shiny app. We strongly advocate for a standard process, and will demonstrate one that has worked really well for us. Start by adding the inputs to the UI. Add the outputs in the UI and server. Then, modify the app layout in the UI. Finally, update the outputs in the server to incorporate user inputs.


```{r}

# create a fluid page interface with a title panel and a sidebar layout
ui <- fluidPage(
        titlePanel(
          "Explore the Gapminder Dataset"
        ),
        # set the theme to "readable"
        theme = shinytheme(
          "readable"
        ),
        sidebarLayout(
          sidebarPanel(
            # create a dropdown menu for selecting a continent
            selectInput(
              "continent",
              "Select Continent",
              # populate the dropdown menu with the distinct values of the "continent" column in the gapminder dataset
              choices = gapminder %>% 
                distinct(continent),
              # set the default selection to "Africa"
              selected = "Africa"
            ),
            # create a slider for selecting a year
            sliderInput(
              "year",
              "Select Year",
              # set the minimum year to 1952
              min = 1952,
              # set the maximum year to 2007
              max = 2007,
              # set the default year to 1952
              value = 1952,
              # set the step size to 5
              step = 5
            )
          ),
          mainPanel(
            # create a tabset panel with two tabs: "Chart" and "Table"
            tabsetPanel(
              tabPanel(
                "Chart",
                # create a plotly output in the "Chart" tab
                plotly::plotlyOutput("plot")
              ),
              tabPanel(
                "Table",
                # create a DT output in the "Table" tab
                DT::DTOutput("table")
              )
            )
          )
        )
      )

# create the server
server <- function(input, output, session){
  # create a function to generate the data to be plotted or displayed in the table
  generate_data <- function(){
    # filter the gapminder data by the selected continent and year
    gapminder %>% 
      filter(continent == input$continent) %>% 
      filter(year == input$year)
  }
  # create a function to generate the plot
  plot_function <- function(){
    # use the generate_data function to get the data to be plotted
    generate_data() %>%
      # create a scatter plot using ggplot
      ggplot(
        # map the x-axis to the "gdp_percap" column, y-axis to the "life_exp" column, 
        # color to the "country" column, and size to the "pop" column
        aes(
          x = gdp_percap,
          y = life_exp,
          color = country,
          size = pop
        )
      ) +
      # add points to the plot
      geom_point() +
      # use the "few" theme from the ggthemes package
      ggthemes::theme_few()+
      # add labels and a caption to the plot
      labs(
        title = "Relationship between Life Expectancy and GDP per Capita",
        subtitle = "Analysis of the gapminder Dataset",
        caption = "Data Source::Gapminder Library",
        x = "GDP per Capita",
        y = "Life Expectancy"
      )+
      # customize the plot background and remove the legend
      theme(
        plot.background = element_rect(fill = "gray90"),
        legend.position = "none"
      )
  }
  # create a plotly plot using the plot_function
  output$plot <- plotly::renderPlotly({
    plot_function()
  })
  # create a DT table using the generate_data function
  output$table <- DT::renderDT({
    generate_data()
  })
}

# run the Shiny app with the defined UI and server
shinyApp(ui, server)



```




